\documentclass[sigplan, nonacm, natbib = false, review, screen]{acmart}

\usepackage{listings}

\lstset{
  basicstyle=\ttfamily\footnotesize,
  columns=fullflexible,
  keepspaces=true
}

\lstdefinestyle{metl}{
  language=[Objective]Caml,
  morekeywords={eff, handle, end, return, mask, forall, Effect, alias},
  morecomment=[l]{--},
  literate=
    {->}{{$\to$}}2
    {=>}{{$\Rightarrow$}}2
    {()}{{()}}2
}

\lstdefinestyle{frank}{
  language=Haskell,
  morekeywords={interface},
  literate=
    {->}{{$\to$}}2
}

\lstdefinestyle{ocaml}{
  language=[Objective]Caml,
  morekeywords={},
  literate=
    {->}{{$\to$}}2
    {=>}{{$\Rightarrow$}}2
    {()}{{()}}2
}

\lstdefinestyle{koka}{
  morekeywords={fun,rec,handler,forall, with, return, ctl, effect, div, type, fn, console},
  sensitive=true,
  morecomment=[l]{//},
  morestring=[b]",
  literate=
    {->}{{$\to$}}2
    {()}{{()}}2
}

\lstdefinestyle{effekt}{
  morekeywords={def, do, interface, import, type, var, in, region, match, case, try, with, box, resume, at},
  sensitive=true,
  morecomment=[l]{//},
  morestring=[b]",
  literate=
    {->}{{$\to$}}2
    {()}{{()}}2
    {=>}{{$\Rightarrow$}}2
}

\usepackage[backend=bibtex, style=numeric]{biblatex}
\addbibresource{abstract.bib}
\renewcommand*{\bibfont}{\footnotesize}

%% Keywords

\newcommand{\keyw}[1]{{\textbf{{\textsf{#1}}}}}
\newcommand{\Do}{\keyw{do}\;}
\usepackage{xspace}
\newcommand{\Met}{\textsc{Met}\xspace}
\newcommand{\MetHO}{$\Met^\sharp$\xspace}
\newcommand{\MetVar}{$\Met^\epsilon$\xspace}

\usepackage{fontawesome5}
\newcommand{\lock}{\textsf{\scriptsize\faLock}}


%% Margin notes.
%% For submission, comment all the following lines out ...
% \paperwidth    =\dimexpr\paperwidth     + 4cm\relax
% \oddsidemargin =\dimexpr\oddsidemargin  + 2cm\relax
% \evensidemargin=\dimexpr\evensidemargin + 2cm\relax
% \marginparwidth=\dimexpr\marginparwidth + 2.2cm\relax
% \usepackage{todonotes,setspace}
% \newcommand{\todobox}[3]{{\todo[color=#1!20,bordercolor=#1!80]{%
%   \begin{spacing}{0.9}\footnotesize\textbf{#2:} #3\end{spacing}}}}
% \newcommand{\wenhao}[1]{\todobox{blue}{Wenhao}{#1}}
% \newcommand{\aghilas}[1]{\todobox{red}{Aghilas}{#1}}
% \newcommand{\sam}[1]{\todobox{yellow}{Sam}{#1}}
%% ... up to here

\usepackage{mathpartir}

\title{Higher-order fork, modally}

\settopmatter{authorsperrow=3}

\author{Aghilas Y. Boussaa}
\orcid{0009-0005-5954-1711}
\email{aghilas.boussaa@ens.fr}
\affiliation{%
  \institution{École normale supérieure -- PSL}
  \country{France}
}

\author{Wenhao Tang}
\orcid{0009-0000-6589-3821}
\email{wenhao.tang@ed.ac.uk}
\affiliation{%
  \institution{The University of Edinburgh}
  \country{UK}
}

\author{Sam Lindley}
\orcid{0000-0002-1360-4714}
\email{sam.lindley@ed.ac.uk}
\affiliation{%
  \institution{The University of Edinburgh}
  \country{UK}}

\begin{document}
\sloppy
\begin{abstract}
  Effect handlers are a powerful programming construct allowing users
  to define and combine computational effects, such as cooperative
  concurrency.
%
  Effect systems statically track the effects that a program can
  perform.
%
  Traditional effect systems rely on effect variables to provide
  modular type signatures for higher-order functions. Recent work on
  modal effect types (\Met) reduces, and often eliminates altogether,
  the need for effect variables via modalities.
%
  Alas, the standard way to define a higher-order fork operation whose
  child process is parametric in its effects is to extend \Met with
  effect variables.
%
  We propose a small extension to \Met, based on introducing an
  ordering on effects in effect contexts, that enables us to define a
  higher-order fork operation whose child process is parametric in its
  effects, without the need for any effect variables at all.
  %
  % We first illustrate these limitations by implementing cooperative concurrency
  % in four existing languages. We then present a new language based on modal
  % effect types, which we use to implement cooperative concurrency with a
  % higher-order fork.
\end{abstract}
\maketitle
\section{Introduction}
Effect handlers can express many control-flow constructs. One example
is cooperative concurrency~\cite{structuredasync2017}, which can be
modelled using an effect with two operations: \textsf{fork} creates a
new child process; \textsf{yield} pauses the current process.
%
% \sam{The following is not really true. Most real implementations of concurrency in these languages use a higher-order fork effect. What are we actually trying to say here?}
% \aghilas{I wanted to say that we can easily implement the unix style fork in all of these languages}
%
There are two canonical forms of fork operation we might consider.

The first form of fork is a UNIX(-style) fork operation of type
$\textsf{unit} \Rightarrow
\textsf{bool}$~\cite{hillerstrom2021thesis} (we denote operation
signatures with a fat arrow).
%
The continuation receives a boolean for distinguishing the parent and
child in the following code.
\begin{lstlisting}[style=metl]
  if do fork () then print "parent" else print "child"
\end{lstlisting}
%
UNIX fork relies on multi-shot continuations, so is unsuitable for
programming languages that support only single-shot
continuations~\cite{ocamleffect2021, wasmeffect2023}.

The second form of fork is a higher-order fork operation of type
$(\textsf{unit} \to \textsf{unit}) \Rightarrow
\textsf{unit}$~\cite{dolan2015effective, ocamleffect2021}.
%
% By taking an alternative approach, one can in fact implement
% cooperative concurrency with single-shot
% continuations~\cite{dolan2015effective, ocamleffect2021}. In OCaml, an
% alternative fork operation can be given following signature.
% \begin{lstlisting}[style=ocaml]
%   type _ Effect.t += Fork : (unit -> unit) -> unit t
% \end{lstlisting}
% This operation takes as input a thunk of the computation which the new
% task should run. The scheduler maintains a queue of thunks that it
% runs. As OCaml has no static effect system, the type is
% straightforward.
This operation takes a thunk for the child process, which is a
function of type $\textsf{unit}\to\textsf{unit}$.
%
The program corresponding to our previous example is written as
follows with higher-order fork.
\begin{lstlisting}[style=metl]
  do fork (fun () -> print "child");
  print "parent"
\end{lstlisting}

Higher-order fork can be implemented using single-shot
continuations.
%
%% SL: This is more-or-less true, but this observation seems to
%% intricate to include right here.
%
%% However, whereas (a slight generalisation of) UNIX fork
%% can be characterised by a (parameterised) algebraic theory,
%% higher-order fork is not even algebraic. Nonetheless, it is still a
%% practically useful operation.
%
However, higher-order fork is somewhat complicated by the need to keep
track of the effects in its parameter.
%
%% However, being higher-order,  equational
%% theories~\cite{KammarLLMS26}, meanwhile presenting extra challenges
%% for effect systems.
%
Traditional effect systems annotate function arrows with effects, so the
argument type $(\textsf{unit} \to \textsf{unit})$ of \textsf{fork} would
typically be annotated with the effects that the child process might
perform. Annotating arrows with the effects they can perform between square
brackets, a suitable scheduler could have a type along the lines:
\begin{lstlisting}[style=metl]
  forall e . (unit ->[co e, queue (proc e), e] unit)
    ->[co e, queue (proc e), e] unit
\end{lstlisting}
%
where \lstinline[style=metl]{co} is the concurrency effect (admitting
\lstinline[style=metl]{fork} and \lstinline[style=metl]{yield}
operations), the \lstinline[style=metl]{queue} effect abstracts over
the queue of concurrent processes and \lstinline[style=metl]{proc} is
a type of suspended processes.
%
We rely on the effect variable \lstinline[style=metl]{e} to
parameterise over the effects that the child process may perform.

Modal effect types (\Met) offer elegant means of avoiding effect variables for
higher-order functions, but still require effect variables for higher-order
operations such as higher-order fork.
%
We write \MetVar for the extension of \Met with effect variables.
%
In the rest of this extended abstract, we briefly describe
implementations of higher-order fork in existing effect systems (Koka,
Frank, Effekt, and \MetVar), demonstrate why effect variables are
needed, and introduce \MetHO, a small extension to \Met which supports
higher-order fork without effect variables.

\section{Higher-order fork in existing languages}\label{impl}

In this section, we briefly sketch how to implement a cooperative
concurrency effect incorporating a higher-order fork operation in
existing languages with effect handlers and effect systems.
%
Our implementations of schedulers use a parametric queue effect to store thunks.
%
The full implementations are available in the appendix and
\href{https://github.com/4y8/hope26}{online}.

\paragraph{Koka}
Koka has a traditional type-and-effect system where each arrow type holds a
scoped row of the effects the function can
perform~\cite{koka2014,scopedrows2005}.
%
We use an effect variable for the fork operation and to define an
effect-polymorphic type of processes. The same effect variable appears on all
arrows of the scheduler's type.

\paragraph{Frank}
Embodying the observation that effect variables are almost always the
same throughout a signature, the Frank programming language omits
effect variables by default and uses syntactic sugar to automatically
insert them~\cite{frank2020}.
%
This allows implementing concurrency with a higher-order fork
operation without any explicit effect
variables~\cite[Section~6]{frank2020}.
%
However, effect variables still exist during type checking and may
leak into error messages displayed to the programmer.

\paragraph{Effekt}
Effekt tracks effects as capabilities~\cite{effekt2020,
effekt2022}.
% This approach relies on functions being second-class citizens in the language.
%
% This particularity offers two ways to define the cooperative concurrency effect,
% none of which uses a queue effect. On the one hand,
%
Functions in Effekt are second-class citizens by default.
%
In order to make them first-class so that we can use them as arguments
of \textsf{fork}, we have to box them and specify the capabilities
they use explicitly.
%
In order to parameterise over the capabilities that the child process
might use, we must introduce an explicit capability variable, which
is reminiscent of the effect variables in Koka and \MetVar.

\section{Higher-order fork in \MetVar}
\label{sec:met}

Modal effect types (\textsc{Met}) decouple effects from function types
via modalities~\cite{tang2025modal, tang2026modal} and enable type
signatures similarly succinct to Frank's without relying on fragile
syntactic sugar.
%
The central notions of \textsc{Met} are the \emph{effect context}, which
determines the effects provided by the environment, and
\emph{modalities}, which transform the effect context.
%
For example, a \lstinline{map} function has the following type
signature.
%
\begin{lstlisting}[style=metl]
  val map : forall a b. []((a -> b) -> list a -> list b)
\end{lstlisting}
%
The \lstinline{[]} is an \emph{absolute modality} which transforms the ambient effect context by discarding it and replacing it with the empty effect context, since
the definition of \lstinline{map} itself uses no effect.
%
%% SL: May be helpful to say something like this:
%
%% (Intuitively, one can think of this transformation applying to every
%% arrow in the type \lstinline[style = metl]{(a -> b) -> list a -> list b}.)
%
We can apply \lstinline{map} to any effectful function, as long as the
effect context at the call site provides the required effects.
%
We refer to \cite{tang2025modal, tang2026modal} for a full
introduction to \textsc{Met}.

\Met enables a large fragment of effect polymorphic programs in \Met
to be written without using effect variables~\cite{tang2025modal}.
%
However, this fragment does not include programs that use higher-order
effects.
%
To ensure effect safety, operations can only yield values whose
meaning is independent of the ambient effect context, which is not the
case for arrow types, as they may use any effect in the
context. Yielding a function is only possible by first boxing it in an
\emph{absolute} modality that fixes the set of effects it can
perform. To be able to compose cooperative concurrency with any
effect, we cannot fix a set of effects and rely on an extension with effect
variables~\cite[Section~2.10]{tang2025modal}.
\begin{lstlisting}[style=metl]
  eff co (e : Effect)
    = fork : [co e, e](unit -> unit) => unit
    | yield : unit => unit
\end{lstlisting}
We define a recursive polymorphic type of processes.
\begin{lstlisting}[style=metl]
  type proc (e : Effect) =
    Proc of [queue (proc e), e](unit -> unit)
\end{lstlisting}
Then, the scheduler has the following type.
\begin{lstlisting}[style=metl]
  forall (e : Effect) . [queue (proc e), e]
    ([co e, queue (proc e), e](unit -> unit) -> unit)
\end{lstlisting}

Because each arrow has an absolute modality, this signature is similar to the
one in the introduction, and loses the advantages of modal effect
types. Moreover, this implementation relies on a \emph{modality-parameterised
handler}~\cite[Section~3.5]{tang2026modal}. These handlers wrap the continuation
with a user-supplied modality. Here, this allows us to fix the effects the
continuation can perform with an absolute modality.

%% SL: mostly addressed
%
%% \sam{There are two many definitions missing here for this to be
%%   properly understood. What even is a modality? What is a
%%   modality-parameterised handler? A citation to a paper isn't good
%%   enough. At least a brief sentence to describe the intuition is
%%   required.}
%
%% \sam{In the next section some relative modalities appear out of thin
%%   air. But relative modalities aren't described here or there. They
%%   need to be.}
%
%% \sam{Extended abstracts, just like papers, need to be pretty much
%%   self-contained. Citations are good, but they cannot replace
%%   explanations.}

%
% Fork can take a first-class boxed function.
% %
% \begin{lstlisting}[style=effekt]
%   interface Co[A, B] {
%     def yield(): Unit
%     def fork(p: Process[A, B]): Unit }

%   type Proc[A, B] =
%     ({c: A => B} => (() => Unit / Co[A, B] at {c})) at {}
% \end{lstlisting}
% To perform any effect other than concurrency, processes have to use the argument
% \texttt{\small c}.
% %

\section{Higher-order fork in \MetHO}


%% SL: mostly resolved
%
%% \sam{This section isn't really readable. Certainly an example would
%%   help - as Wenhao says.}

As we have seen, in order to implement a higher-order fork effect that
allows the child process to perform any effect, we have to use effect
variables either explicitly or implicitly in existing effect systems.
%% SL: space-saving
%
%% such as Koka, Frank, Effekt, and \textsc{Met} .
%
This is particularly unsatisfactory for \textsc{Met}, as \textsc{Met}
is designed to avoid effect variables.

We observe that the main challenge is that when calling a higher-order
fork in \Met, we do not know what the effect context for the argument
(the child process) should be, as it depends on the effect context of
the handler of fork.
%
% Effect tracking of \textsc{Met} relies on tracking switches between
% effect contexts, while higher-order effects such as the higher-order
% fork allows us to capture a closure as an operation argument and later
% use it at the handler side, which enables switching between effect
% contexts in an opaque way.
%
As a result, we must fully specify which effects may be performed by
the child process via an absolute modality as in
Section~\ref{sec:met}, and consequently, we use effect variables to
parameterise over over the effects that the child process might perform.
%
% As a result, we must fully specify which effects might be performed by
% the child process with an absolute modality as in
% Section~\ref{sec:met}, and consequently, we must use effect
% variables to be parameterised over these effects.

We propose \MetHO, a simple solution to avoid effect variables.
%
Our key idea is to impose an order on effects in effect contexts
reflecting the order of handlers in the evaluation context.
%
When we call a higher-order operation such as fork, the
effects available when the child process is called are exactly those
that follow the \textsf{co} effect in the effect context.
%
%% SL: mostly resolved
%
%% \wenhao{Would be helpful to give a concrete example here}
%
% The unsatisfactory solution in \textsc{Metl} motivates the search for new ways
% to extend the core calculus with support for polymorphic higher-order effects. We have
% implemented a promising approach: relax kind restrictions on operations’
% operands, but restrict subeffecting in presence of non-absolute types. These
% conditions should ensure that a non-absolute value arrives to an effect context
% more permessive than the one it comes from. Appendix~\ref{theory} presents the
% essential rules.
%
The cooperative concurrency effect is now defined as follows.
\begin{lstlisting}[style=metl]
  eff# co = fork : (unit -> unit) => unit
          | yield : unit => unit
\end{lstlisting}
The hash (\texttt{\small \#}) indicates that \textsf{co} is a higher
order-effect, allowing \textsf{fork} to take as input a function.
%
%% SL: I think the following is now resolved
%
%% \sam{This description of bang is entirely opaque. Most readers will
%%   have no idea what it is intended to mean.}
%
For example, in presence of a function
\lstinline[style = metl]{f : <queue unit>(unit -> unit) -> unit}, the typing
rules reject the unsafe program below, which would allow the queue effect to
leak~\cite[Section~2.10]{tang2025modal}.
\begin{lstlisting}[style = metl,escapechar=\%]]
  handle f %\underline{(fun () -> do fork (fun () -> do enqueue ()))}% with
    | fork p _ => p -- we omit the other clauses
\end{lstlisting}
The modality \lstinline[style = metl]{<queue unit>} is
% \emph{relative}. Instead of replacing the ambient effect context,
% relative modalities extend it. Here, the modality adds the queue
% effect in front of the effect context of the underlined thunk. The
% ordering condition ensures that \textsf{fork} cannot be performed in
% this context, where it is not first.
% %
%
a \emph{relative} modality, which extends the ambient effect context
with \lstinline{queue unit}.
%
As a result, the effect context for the underlined thunk would be
\lstinline[style = metl]{queue unit, co}.
%
From this ordering, \MetHO knows that the queue effect would not be
available when the child process is called at the handler of
\textsf{fork}, rejecting this program.

The signature of \textsf{fork} does not mention the \textsf{co}
effect, because the thunk will have the same ambient effect context as
the computation that yielded it, which contains effect \textsf{co}.
%
We can further eliminate the \lstinline{proc} datatype and just use
plain functions.
%
The scheduler does not need a modality-parameterised handler and
has the following type.
\begin{lstlisting}[style=metl]
  [queue (unit -> unit)](<co>(unit -> unit) -> unit)
\end{lstlisting}

In summary, \MetHO offers a structural way to determine the set of effects
the operands an operation can perform. In the case of higher-order
fork, this eliminates the need for effect variables entirely. Future
work includes: extending modalities to also act on effects to increase
the flexibility of \MetHO, and inserting masks instead of requiring
higher-order operations to come from the first effect in the context.
%% Another difference with the implementations of Section~\ref{impl} is
%% the absence of masking.
%
%% SL: I think masking has now disappeared so this comment is resolved
%
%% \sam{This is perplexing. We haven't said anything about
%%   masking. Perhaps this is because of shipping code to the
%%   appendix. In any case we must explain masking if we are to talk
%%   about it. Ideally we'd try to avoid talking about it in the body of
%%   the paper though.}
%
%% The semantics wraps the operands of higher-order effects with the same
%% masks as the yielding evaluation context. The first call to the
%% scheduler still masks the \textsf{queue} effect.
%% \begin{lstlisting}[style=metl]
%%   val full_schedule : [](<co>(unit -> unit) -> unit)
%%   let full_schedule m =
%%     queue { schedule { mask<queue> (m!) } }
%% \end{lstlisting}

%% SL: This is true, but not particularly interesting
%
%% For now, we lack a proof of safety and an expressivity result relating this
%% system with the one with effect variables.

%% SL: more-or-less resolved

%% \sam{This is an extremely abrupt end. It would be good to give a bit
%%   more detail on how this actually works. It would also be nice to
%%   finish with some kind of convincing conclusion, takeaway message, or
%%   exciting proposal for future work.}

%% \wenhao{Would it make sense to modify the implementation to automatically
%% wrap the argument when calling higher-order fork, instead of requiring
%% the effect to be the first?}

\printbibliography
\appendix

\section{Koka}
\lstinputlisting[style=koka]{app/koka.kk}
\section{Frank}
\lstinputlisting[style=frank]{app/frank.fk}
\section{\MetVar}
\lstinputlisting[style=metl]{app/metl-eff-var.met}

\section{Effekt}
\subsection{Boxed functions}
\lstinputlisting[style = effekt]{app/effekt-boxes.effekt}
\subsection{Bidirectional effects}
Effekt supports bidirectional effects which can be used to implement
some form of higher-order effects.
%
With bidirectional effects, fork can take as input second-class
\emph{blocks}.
%
\begin{lstlisting}[style = effekt]
  interface Co {
    def yield(): Unit
    def abort(): Nothing
    def forkOp{ p: () => Unit }: Unit }
\end{lstlisting}
When handling \texttt{\small forkOp}, the continuation expects as argument a
function that takes \texttt{\small p} as input. The scheduler resumes the
continuation twice: to enqueue \texttt{\small p} and to continue the
computation.
% The type checker does not allow enqueuing \texttt{\small p} from the continuation's argument
To continue the parent computation only once,
the \texttt{\small fork} function wraps each fork with an \texttt{\small abort}
operation.
\begin{lstlisting}[style = effekt]
  def fork { p: => Unit / Co }: Unit / Co =
    do forkOp { p(); do abort() }
\end{lstlisting}

The full implementation is as follows.

\lstinputlisting[style = effekt]{app/effekt-bieff-with-abort.effekt}

\subsection{Additional comments}
The version with bidirectional effects resumes the continuation twice. Effekt's
type system reject the following handler for \texttt{\small forkOp}, which would
only require resuming once, because the capability \texttt{\small child} is
incompatible with the queue.
\begin{lstlisting}[style = effekt]
  def forkOp() =
    resume { {child: () => Unit} =>
      queue.pushFront(box child);
      () }
\end{lstlisting}
Such a handler, would remove the need for the wrapper.

We have not managed to use a queue effect to store our thunks. To do so, we
would must store in a queue a thunk that captures the queue capability
itself. However, we could not implement such a queue while staying in Effekt's
lexical requirements.
\section{\MetHO}
% \subsection{Implementation}
\lstinputlisting[style=metl]{app/metl-ho-eff.met}


%% SL: I think this is helpful, so I've restored it for now...

% WT: Given that we have more explanations in the main text, I don't
% think we need this appendix any more.
%\iffalse
\subsection{Theory}\label{theory}
%
We here spell out the changes to the formal rules of
$\Met(\mathcal{R}_{\textsf{scp}})$~\cite{tang2026modal} necessary to support our
extension. We only consider here a simplified version without local labels and
modality-parameterised handlers.
%
The syntax annotates each effect type with an \emph{accidental}: a flat
signature $A\Rightarrow^\flat B$ has operands of absolute types, and a
sharp signature $A\Rightarrow^\sharp B$ has no restriction. Effect
contexts and extensions are scoped rows of labels that a context map to
signatures.
\begin{center}
\begin{tabular}{ l r c l }
  Accidentals & $Y$ & ::= & $\flat\mid\sharp$\\
  Extensions & $D$ & ::= & $\cdot\mid\ell,D$\\
  Effect Contexts & $E,F$ & ::= & $\cdot\mid\ell,E$\\
  Label Contexts & $\Sigma$ & ::= &
  $\cdot\mid\ell:A\Rightarrow^Y B,\Sigma$\\
\end{tabular}
\end{center}
The extension equivalence rules fix the position of sharp signatures. Effect
context equivalence obeys the same rules. All of the following rules suppose
the existence of a global label context $\Sigma$.
\begin{mathpar}
  \inferrule{\\}{\cdot\equiv\cdot}

  \inferrule{D_1\equiv D_2\\ D_2\equiv D_3}{D_1\equiv D_3}

  \inferrule{D_1\equiv D_2}{\ell, D_1\equiv \ell, D_2}

  \inferrule{\ell\neq \ell'\\ \ell:\flat\\ \ell':\flat}
    {\ell, \ell', D \equiv \ell', \ell, D}
\end{mathpar}
Subeffecting of extensions allows moving sharp labels backward.
\begin{mathpar}
  \inferrule{\\}{\cdot\leqslant\cdot}

  \inferrule{D_1\leqslant D_2\\ D_2\leqslant D_3}{D_1\leqslant D_3}

  \inferrule{D_1\leqslant D_2}{\ell, D_1\leqslant \ell, D_2}

  \inferrule{\ell\neq \ell'\\ \ell:\flat}
    {\ell, \ell', D \leqslant \ell', \ell, D}
\end{mathpar}
For subeffecting, effect contexts follow the same rules as extensions with the
addition of $\cdot\leqslant E$.

The rules for $\keyw{do}$ ensure that a sharp operation is available only if it appears as the first operation in the ambient effect context.
\begin{mathpar}
  \inferrule[\textsc{T-DoFlat}]
    {E = D,\ell,F\\
     \ell\not\in D\\
     \Sigma\ni\ell:A\Rightarrow^\flat B\\
     \Gamma\vdash N:A \;@\; E}
    {\Gamma\vdash\Do\ell\; N:B \;@\; E}

  \inferrule[\textsc{T-DoSharp}]
    {\Sigma\ni\ell:A\Rightarrow^\sharp B\\
     \Gamma\vdash N:A \;@\; \ell,E}
    {\Gamma\vdash\Do\ell\; N:B \;@\; \ell,E}
\end{mathpar}

To simplify the rules, we restrict each handler to a single operation. The rule
\textsc{T-HandleSharp} wraps the operands of sharp operations with extension
modalities.
\begin{mathpar}
  \inferrule[\textsc{T-HandleSharp}]
    {\Gamma,\lock_{{\langle|\ell\rangle}_E}\vdash M:A\;@\;\ell,F\\
     \Sigma\ni\ell:A'\Rightarrow^\sharp B'\\
     \Gamma, x:\langle\ell\rangle A\vdash N:B\; @\; F\\
     \Gamma, p:\langle\ell\rangle A', r:\langle\ell\rangle B'\to B\vdash
     N':B\; @\; F
    }
    {\keyw{handle } M\keyw{ with }
     \{\keyw{return }x\mapsto N,\ell\;p\;r\mapsto N'\}
     : B\; @\; F}
\end{mathpar}
Handling flat operation works the same way as usual in \textsc{Met}.
%\fi
\end{document}

% LocalWords:  bool boolean equational Koka Effekt modally enqueue
% LocalWords:  dequeue Koka's positivity ctl proc koka qproc forall
% LocalWords:  effectful expressivity Metl metl effekt forkOp DoFlat
% LocalWords:  Effekt's pushFront Subeffecting subeffecting DoSharp
% LocalWords:  HandleSharp scheduler's scp
